FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2017-10-12 NewVersion/sections/Formalization.tex
Section 3

\section{Formalization}~\label{sec:formalization}
In this section, we present the formal model called \MIM{} (\emph{\textbf{F}eatherweight \textbf{H}ierarchical \textbf{J}ava}), following similar style of 
Featherweight Java~\cite{Igarashi01FJ}. \MIM{} is a minimal core calculus that formalizes the core concept of hierarchical dispatch and overriding. The syntax, typing rules,
small-step semantics and theorems for type soundness are presented.

@@

\textsc{(T-Invk)} is the typing rule for method invocation.
Naturally, the receiver and the arguments are required to be well-typed.
$\mbody$ is our key function for method lookup, based on the
hierarchical dispatching algorithm. The formal definition will be introduced in
Section~\ref{sec:auxdefs}. To overview,
the hierarchical invocation $\mbody(m, I, J)$ means ``finding most specific $m$
above $I$ and along path $J$'', and usually $I$ denotes the dynamic type while $J$
denotes the static type. In \textsc{(T-Invk)}, as the compilation should not be aware
of the dynamic type, it only requires that invoking $m$ is valid for the static type of the
receiver. The result of $\mbody$ contains the interface that provides the most specific implementation,
the parameters and the return type. We use underscore for the return expression, to imply that an empty return expression
from an abstract method is also acceptable.
Note that we accept implicit upcasts on the arguments.
% \textsc{(T-PathInvk)} is the typing judgement for a path invocation. Besides the conditions of \textsc{(T-Invk)}, \textsc{(T-PathInvk)} requires the type of receiver to be the subtype of the specified path type. 
% and \textsc{(T-SuperInvk)} checks if the enclosing type directly extends the specified super type.

\textsc{(T-New)} is the typing rule for object creation $\new{I}$. The auxiliary function $\canInstantiate(I)$ checks whether interface $I$ can be instantiated or not, formalized in Section~\ref{sec:auxdefs}. Since triangle inheritance accepts conflicting branches to coexist, the check requires that the most specific method is concrete for each method on each branch.

\textsc{(T-Method)} is more interesting since a method can either be an original method or a hierarchical overriding, though
they share the same syntax and method typing rule. $\mostSpecific(m, I, J)$ is a fundamental function from Section~\ref{sec:auxdefs},
used to find ``the most specific interfaces that are above $I$ and along path $J$, and originally defines $m$''. By ``most specific interfaces'',
it implies that the inherited super types are excluded. Thus the condition $\mostSpecific(m, I, J) = \{J\}$ indicates a characteristic of a hierarchical overriding: it must override an original method; the overriding is direct and there does not exist any other original $m$ in between.
Then $\mbody(m, J, J)$ provides the type of the original method, so hierarchical overriding has to preserve the type. Finally the return expression
needs to be type-checked, and there is also an implicit cast to the declared return type. For the definition of an original method, $I$ equals $J$ and the rule is straightforward. \textsc{(T-AbsMethod)} is a similar rule but works on abstract method declarations.

@@

We present the type soundness of the model by a few theorems below, following the standard technique of
subject reduction and progress proposed by Wright et al.~\cite{Wright1994}. The proof, together with some lemmas, are
in Appendix. Type soundness states that if an expression is well-typed, then after many reduction
steps it must reduce to a value, and its annotation is the same as the static type of the original expression.

>>>>

\section{Formalization}~\label{sec:formalization}
In this section, we present a formal model called \MIM{} (\emph{\textbf{F}eatherweight \textbf{H}ierarchical \textbf{J}ava}), following similar style of 
Featherweight Java~\cite{Igarashi01FJ}. \MIM{} is a minimal core calculus that formalizes the core concept of hierarchical dispatch and overriding. The syntax, typing rules,
small-step semantics and theorems for type soundness are presented.

@@

\textsc{(T-Invk)} is the typing rule for method invocation.
Naturally, the receiver and the arguments are required to be well-typed.
$\mbody$ is our key function for method lookup, based on the
hierarchical dispatching algorithm. 
The hierarchical invocation $\mbody(m, I, J)$ means ``finding most specific $m$
above $I$ and along path $J$'', and usually $I$ denotes the dynamic type while $J$
denotes the static type. 
The formal definition of $\mbody(m, I, J)$ will be introduced in
Section~\ref{sec:auxdefs}.
In \textsc{(T-Invk)}, as the compilation should not be aware
of the dynamic type, it only requires that invoking $m$ is valid for the static type of the
receiver. The result of $\mbody$ contains the interface that provides the most specific implementation,
the parameters and the return type. We use underscore for the return expression, to imply that an empty return expression
from an abstract method is also acceptable.
Note that we accept implicit upcasts on the arguments.
% \textsc{(T-PathInvk)} is the typing judgement for a path invocation. Besides the conditions of \textsc{(T-Invk)}, \textsc{(T-PathInvk)} requires the type of receiver to be the subtype of the specified path type. 
% and \textsc{(T-SuperInvk)} checks if the enclosing type directly extends the specified super type.

\textsc{(T-New)} is the typing rule for object creation $\new{I}$. The
auxiliary function $\canInstantiate(I)$ checks whether interface $I$
can be instantiated or not, and is formalized in Section~\ref{sec:auxdefs}. Since triangle inheritance accepts conflicting branches to coexist, the check requires that the most specific method is concrete for each method on each branch.

\textsc{(T-Method)} is more interesting since a method can either be an original method or a hierarchical overriding, though
they share the same syntax and method typing rule. $\mostSpecific(m, I, J)$ is a fundamental function,
used to find ``the most specific interfaces that are above $I$ and
along path $J$, and originally defines $m$'' (see
Section~\ref{sec:auxdefs} for full definition).
By ``most specific interfaces'',
it implies that the inherited super types are excluded. Thus the condition $\mostSpecific(m, I, J) = \{J\}$ indicates a characteristic of a hierarchical overriding: it must override an original method; the overriding is direct and there does not exist any other original $m$ in between.
Then $\mbody(m, J, J)$ provides the type of the original method, so hierarchical overriding has to preserve the type. Finally the return expression
needs to be type-checked, and there is also an implicit cast to the declared return type. For the definition of an original method, $I$ equals $J$ and the rule is straightforward. \textsc{(T-AbsMethod)} is a similar rule but works on abstract method declarations.

@@

We present the type soundness of the model by a few theorems below, following the standard technique of
subject reduction and progress proposed by Wright and Felleisen~\cite{Wright1994}. The proof, together with some lemmas, are
in Appendix. Type soundness states that if an expression is well-typed, then after many reduction
steps it must reduce to a value, and its annotation is the same as the static type of the original expression.
